Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-06-05 Manuscript/manuscript.Rmd
edits

Current methods used to predict TAD boundaries rely on the use of classification algorithms to draw associations between the spatial relationship of functional genomic elements and regions that include TAD boundaries. However, this is a multi-faceted probelem, and many methods ignore key characteristics [?? which ones ??] of domain data when building models and reporting performances. We aim to address these characteristics in order to create a more unified approach toward identifying the genomic annotations most predictive of TAD boundaries.

A major ignored aspect of TAD boundary prediction is the introduction of heavily imbalanced classes resulting from the number of boundaries vs. nonboundary regions. In a classification setting, when one is attempting to discriminate between minority (regions with TAD boundaries) and majority classes (regions without TAD boundaries), the class imbalance issue will contribute to small imbalance ratios. The imbalance ratio is defined as the ratio of minority to majority classes [@Orriols-Puig2008]. Such heavily imbalanced datasets are known to adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class; @chen2013influence; @jeni2013facing] [?? This sentence is very difficult to understand. Don't see connection between the bias and the error rate ??].

In a study published in 2016 [@mourad2016computational], an L1-regularized multiple logistic regression model was proposed using transcription factor binding sites on GM12878 to discriminate between regions with and without TADs, and to identify the most predictive factors associated with boundaries. However, the class imbalance issue was not addressed. Even worse, when performing cross-validation, [@mourad2016computational] used the misclassification error rate of the model to establish the optimal L1-regularization lambda. They also reported ROC (receiver operating characteristic) curves and subsequent AUC (area under the curve) as measures of their model's performance. This is problematic in an imbalanced setting. Such heavily imbalanced datasets are known to adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class; @chen2013influence; @jeni2013facing] [?? This is the second mention of "biased towards the majority class to minimize the overall error rate" without strong explanation of what this means ??]. For AUC, when the prevalence of an event is low--that is, a defined genomic region truly containing a TAD boundary--mean probabilities are biased towards the majority class. This results in the area of interest under an ROC curve to be compressed toward a small corner in the ROC space [@prati2011survey; @wei2013role]. While not directly affecting the value of the AUROC, this can have severe consequences on the interpretability of the model's performance [?? How so? does it inflate it or deflate it or what ??]. Consequently, careful consideration must be taken when deciding which performance metric(s) to use for evaluation of predictive models, especially in the case of imbalanced data.

A better threshold-free measure to use when dealing with imbalanced data is the area under the precision-recall curve (AUPRC).  The interpretation of the AUPRC of a model is similar to that of the AUROC. However, where as the baseline model performance according to the AUROC is given by 0.50, the baseline model performance according to the AUPRC is given by the ratio of true positives over the total number of cases [?? Need to identify what you mean by "total number of cases." Is it just true positive divided by total number of possible boundaries ??]. This enables a more realistic benchmark to evaluate a model on as this ratio will be lower than 0.5 for a test data set with imbalanced classes [@saito2015precision]. Another threshold-specific measure to consider is a composite metric involving all quadrants of the confusion matrix, such as Matthews Correlation Coefficient (MCC) [@jeni2013facing; @chicco2017ten] [?? MCC actually has all of the same components as AUC. AUC is a function of sensitivity and specificity which combined contain all quandrants ??]. The MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications [@brodersen2010balanced]. A value of 1 represents a perfect prediction, while a value of -1 represents total disagreement between prediction and observation. A value of 0 indicates no better than random prediction.

We can account for imbalance at the data level in addition to at the evaluation level [?? I rewrote this sentence. Was very wordy and difficult to follow ??]. Data level solutions mostly constitute as [?? I'd change "constiute as", unorthodox word choice ??] re-sampling techniques, in which random sampling is used to either replicate or remove data points in order to create a more balanced dataset [@chawla2004special; @chawla2009data]. It has been shown that quite simple re-sampling techniques can drastically improve model performance in the face of imbalanced data [@japkowicz2002class; @dubey2014analysis]. [?? It feels like data level should come before evaluation level ??]

Two studies published in 2015 and 2017 respectively [@huang2015predicting; @hong2017computational] used re-sampling techniques prior to modeling, and both implemented random under-sampling to balance classes. However, given the large disparity between the number of genomic regions labeled as containing a TAD boundary or not, performing random under-sampling alone is likely to be unstable and introduce bias. Also, it is unclear how other data level solutions compare to simple random under-sampling such as random over-sampling or a weighted combination of both random under- and over-sampling together. Equally concerning is how the test set is treated in scenarios of re-sampling. It was found that [@huang2015predicting] performed re-sampling prior to splitting the data into training and testing. However, this is problematic because re-sampling prior to splitting can further introduce bias into the testing set, as it is no longer an accurate representation of real-world data [@wei2013role].

@@

We analyzed [??? Boring. What problem you are solving?] publically available Hi-C data for the GM12878 and K562 cell lines provided by Rao et al. [@rao20143d]. Topologically associating domains were identified using the Arrowhead algorithm [@durand2016juicebox] obtained at 10 kb, 25 kb, 50 kb, and 100 kb resolutions. We then established a unified approach toward preparing cell line and resolution specific TAD boundary data [??? This is problem definition, should be done early on, not here] for predictive modelling purposes. Specifically, we created a response variable that indicated whether a genomic bin overlapped with a flanked TAD boundary ($Y=1$), or not ($Y=0$, Figure 1). <!--The unit width of each bin on the linear genome corresponded to the resolution of the cell line specific contact matrix on which domains were called. Moreover, TAD boundaries were flanked on either side by the same unit length as the genomic bins to account for uncertainty in boundary location.-->

@@

## Three types of associations between TAD boundaries and epigenomic features

@@

To establish the feature space for classification modeling, we constructed three different predictor types. Counts of overlaps (**overlap count**, OC) is the most commonly used type of association between genomic elements (Figure 2A). **Overlap percent** (OP) represents a more precise measure of associations as it considers the proportion of a bin overlapped by a genomic annotation (Figure 2B). **Distance** between the centers of bins and genomic annotations quantifying the spatial relationship between genomic elements is the least common type of association (Figure 2C). Log2-transformation of distance was used as a normalization technique in order to account for the skewness present in the distributions of distances (Supplementary Figures 1-2). These types of genomic associations provide a granular way to investigate classification performance in class imbalance settings.

>>>>

Current methods used to predict TAD boundaries rely on the use of classification algorithms to draw associations between functional annotations and regions that include TAD boundaries. However, this is a multi-faceted probelem. Many methods ignore key characteristics [?? which ones ??; !!!editted] of 3D genomic data including sparsity of TAD boundaries as well as suitably describing the spatial relationship between annotations and TAD boundaries.

A major ignored aspect of TAD boundary prediction is the introduction of heavily imbalanced classes resulting from the number of boundaries vs. nonboundary regions. In a classification setting, when one is attempting to discriminate between minority (regions with TAD boundaries) and majority classes (regions without TAD boundaries), the class imbalance issue will contribute to small imbalance ratios. The imbalance ratio is defined as the ratio of minority to majority classes [@Orriols-Puig2008]. Such heavily imbalanced datasets are known to adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class; @chen2013influence; @jeni2013facing] [?? This sentence is very difficult to understand. Don't see connection between the bias and the error rate ??; !!!].

We can account for imbalance at the data level in addition to at the evaluation level [?? I rewrote this sentence. Was very wordy and difficult to follow ??]. Data level solutions mostly involve re-sampling techniques, in which random sampling is used to either replicate or remove data points in order to create a more balanced dataset [@chawla2004special; @chawla2009data]. It has been shown that quite simple re-sampling techniques can drastically improve model performance in the face of imbalanced data [@japkowicz2002class; @dubey2014analysis]. Evaluation level solutions include choosing suitable performance metrics for evaluating models. <!--These solutions are necessary given that the test set, used to validate re-sampled training sets, is itself imbalanced.--> It has been shown that common metrics such as accuracy and AUROC (area under the receiver operating characteristic curve) are insensitive to class imbalance [@prati2011survey; @wei2013role]. A better measure to consider would be a composite metric involving all quadrants of the confusion matrix such as Matthews Correlation Coefficient (MCC). The suitability of MCC when evaluating models in imbalanced settings has researched previously [@boughorbel2017optimal; @brodersen2010balanced]. Defined, the MCC is the correlation coefficient between the observed and predicted binary classifications. A value of 1 represents a perfect prediction, while a value of -1 represents total disagreement between prediction and observation. A value of 0 indicates no better than random prediction.

<!--
In a study published in 2016 [@mourad2016computational], an L1-regularized multiple logistic regression model was proposed using transcription factor binding sites on GM12878 to discriminate between regions with and without TADs, and to identify the most predictive factors associated with boundaries. The class imbalance issue was not addressed. Even worse, when performing cross-validation, [@mourad2016computational] used the misclassification error rate of the model to establish the optimal L1-regularization lambda. They also reported ROC (receiver operating characteristic) curves and subsequent AUC (area under the curve) as measures of their model's performance. 
-->

<!--
A better threshold-free measure to use when dealing with imbalanced data is the area under the precision-recall curve (AUPRC).  The interpretation of the AUPRC of a model is similar to that of the AUROC. However, where as the baseline model performance according to the AUROC is given by 0.50, the baseline model performance according to the AUPRC is given by the ratio of true positives over the total number of cases [?? Need to identify what you mean by "total number of cases." Is it just true positive divided by total number of possible boundaries ??]. This enables a more realistic benchmark to evaluate a model on as this ratio will be lower than 0.5 for a test data set with imbalanced classes [@saito2015precision]. Another threshold-specific measure to consider is a composite metric involving all quadrants of the confusion matrix, such as Matthews Correlation Coefficient (MCC) [@jeni2013facing; @chicco2017ten] [?? MCC actually has all of the same components as AUC. AUC is a function of sensitivity and specificity which combined contain all quandrants ??]. The MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications [@brodersen2010balanced]. A value of 1 represents a perfect prediction, while a value of -1 represents total disagreement between prediction and observation. A value of 0 indicates no better than random prediction.
-->

<!--
Two studies published in 2015 and 2017 respectively [@huang2015predicting; @hong2017computational] used re-sampling techniques prior to modeling, and both implemented random under-sampling to balance classes. However, given the large disparity between the number of genomic regions labeled as containing a TAD boundary or not, performing random under-sampling alone is likely to be unstable and introduce bias. Also, it is unclear how other data level solutions compare to simple random under-sampling such as random over-sampling or a weighted combination of both random under- and over-sampling together. Equally concerning is how the test set is treated in scenarios of re-sampling. It was found that [@huang2015predicting] performed re-sampling prior to splitting the data into training and testing. However, this is problematic because re-sampling prior to splitting can further introduce bias into the testing set, as it is no longer an accurate representation of real-world data [@wei2013role]. 
-->

@@

We analyzed [??? Boring. What problem you are solving?] publically available Hi-C data for the GM12878 and K562 cell lines provided by Rao et al. [@rao20143d]. Topologically associating domains were identified using the Arrowhead algorithm [@durand2016juicebox] obtained at 10 kb, 25 kb, 50 kb, and 100 kb resolutions. We then established a unified approach toward preparing cell line and resolution specific TAD boundary data [??? This is problem definition, should be done early on, not here] for predictive modelling purposes. Specifically, we created a response variable that indicated whether a genomic bin overlapped with a flanked TAD boundary ($Y=1$), or not ($Y=0$) (Figure 1). <!--The unit width of each bin on the linear genome corresponded to the resolution of the cell line specific contact matrix on which domains were called. Moreover, TAD boundaries were flanked on either side by the same unit length as the genomic bins to account for uncertainty in boundary location.-->

@@

## DEVELOPING THREE TYPES OF ASSOCIATIONS BETWEEN TAD BOUNDARIES AND EPIGENOMIC FEATURES

@@

To establish the feature space used for classification, we constructed three different predictor types. Counts of overlaps (**overlap count**, OC) is the most commonly used type of association between genomic elements (Figure 2A). Percents of overlaps (**overlap percent**, OP) represents a more precise measure of association as it considers the proportion of a bin overlapped by a genomic annotation (Figure 2B). **Distance** between the centers of bins and genomic annotations quantifying the spatial relationship between genomic elements is the least common type of association (Figure 2C). Log2-transformation of distance was used as a normalization technique in order to account for the skewness present in the distributions of distances (Supplementary Figures 1-2). These types of genomic associations provide a granular way to investigate classification performance in class imbalance settings.
